A Note on Convergence Diagnostics for Multiple Imputation Algorithms

### 2020-01-15 2.ResearchReport/1.SimulationStudy/Writeup/NoteConvergence.Rnw

@@

%% another package (only for this demo article)
% \usepackage{framed}
\usepackage{multirow, amsmath, tikz}
\usetikzlibrary{fit, positioning}

@@

%% - \Abstract{} almost as usual
\Abstract{
This note investigates how algorithmic convergence of multiple imputation procedures---methodology to circumvent the ubiquitous problem of missing data---may be diagnosed. We conclude that potential scale reduction factor $\widehat{R}$ and autocorrelation may be appropriate convergence diagnostics, but their performance differs from conventional applications to iterative algorithmic procedures. These results will be implemented into an interactive framework for the evaluation of multiple imputation methods that is under development as Research Master's thesis. %\textbf{TO DO still: check worst lin pred ref, check ref list, add legend to sim diags, check table and fig captions (abbrevations), references to tables, figs, appendix}
%%%%%%%%%%%%%
%The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI). With MI, missing datapoints are `filled in' (i.e. imputed) several times, and statistical inference is performed on each imputed data set. The validity of inferences may be threatened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software. This note investigates the performance of two convergence diagnostics (\citep[$\widehat{R}$,][]{gelm92}; \citep[autocorrelation,][]{scha97}) through model-based simulation. To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred. Each of these 100 simulation conditions is replicated 1000 times. The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}, since no benchmark diagnostic is available. This simulation study shows that three iterations may be sufficient to obtain ... \textbf{This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{arXiv}. That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as an online appendix to the technical paper on ShinyMICE.}  
%%%%%%%%%%%%
% This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary}''. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}. That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
  }

@@

%% -- Introduction -------------------------------------------------------------

\section{Introduction} \label{sec:intro} % curly brackets can handle typesetting

@@

To obtain valid inferences with MI, the variability between imputations should be properly represented \citep{rubin87, buur18}. If this variability is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. Over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power. Since both of these situations are undesirable, imputations and their variability should be evaluated. Evaluation measures, however, are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}. %\textbf{ADD sentence to link eval measures to convergence!}
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods. These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation, which will aid applied researchers in drawing valid inference from incomplete datasets.

This note provides the theoretical foundation towards the diagnostic evaluation of %the convergence of 
multiple imputation algorithms. For reasons of brevity, we only focus on the MI algorithm implemented in \pkg{mice} \citep{mice}. 
The convergence properties of this MI algorithm are investigated through model-based simulation\footnotemark.%The results of this simulation study are guidelines for assessing convergence of MI algorithms. These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. 
\footnotetext{All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.} 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{autocorrelation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.
We will evaluate how convergence of the MI algorithm can be diagnosed.

% for diagnostics for evalua one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm.  %The implementation limits itself to the MI algorithm 
%More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}.

%The research question that will be addressed is: %. a measure that is vital in evaluating one vital measure for evaluating MI procedures--an algorithmic convergence diagnostic. This note specifically investigates one vital evaluation measure: how to diagnose convergence of a multiple imputation algorithm. 
%This note focusses on one measure that is vital for evaluating the validity of multiply imputed data: a convergence diagnostic. of the MI algorithm. 
%Therefore, the aim of this research project is to investigate which measures are vital for . how the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE). This note addresses the question: 
%\emph{`How to diagnose convergence of the multiple imputation algorithm MICE?'.}

%% -- Features ---------------------------------------------------------------
\subsection{Terminology} \label{sec:terms}

%The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problems.

This note follows notation and conventions of \pkg{mice} \citep{mice}. % to suit the \pkg{mice} environment. %Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}. 
Basic familiarity with MI methodology is assumed.\footnote{For the theoretical foundation of MI, see \cite{rubin87}. For an accessible and comprehensive introduction to MI from an applied perspective, see e.g. \cite{buur18}.}

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted by $Y_{obs}$, and will be referred to as `incomplete' or `observed' data. The missing part of $Y$ is denoted by $Y_{mis}$. Which parts of $Y$ are missing is determined by the `missingness mechanism'. %The relation between $Y_{obs}$ and $Y_{mis}$ \textbf{REMOVE? Response indicator $R$ shows whether a data value in $Y$ is missing or observed.} The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. 
This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$ \citep{rubin87}.

%In this note, the terms `unobserved' and `missing' data are used interchangeably to refer to $Y_{mis}$. The terms `incomplete' or `observed' data denote $Y_{obs}$. Incomplete data is the starting point of the multiple imputation procedure.

@@

\begin{figure}
\label{fig:steps}
\centering
	\large{\begin{tikzpicture}
	\tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 20mm]
\tikzstyle{connect}=[-latex, thick]
\node[main, fill = white!100] (data) [label=left:Incomplete data] { };
\node[main] (mids) [right=of data] { };
\node[main] (mids2) [above=10mm of mids] { };
\node[main] (mids3) [below=10mm of mids,label=below:Imputed data] { };
\node[main] (mira) [right=of mids] {};
\node[main] (mira2) [above=10mm of mira] { };
\node[main] (mira3) [below=10mm of mira,label=below:Analysis results] { };
\node[main, fill = white!100] (mipo) [right=of mira,label=right:Pooled result] { };
\path (data) edge [connect] (mids)
      (data) edge [connect] (mids2)
      (data) edge [connect] (mids3)
      (mids) edge [connect] (mira)
      (mids2) edge [connect] (mira2)
      (mids3) edge [connect] (mira3)
		  (mira) edge [connect] (mipo)
		  (mira2) edge [connect] (mipo)
		  (mira3) edge [connect] (mipo);
\end{tikzpicture}}
\caption{Scheme of the main steps in multiple imputation \citep[$m = 3$; adapted from][\S~1.4.1]{buur18}. {\footnotesize Missing data in dataset $Y$ is `imputed' (i.e., filled in) $m$ times. The imputed data is combined with the observed data ($Y_{obs}$) to create $m$ completed datasets. On each completed dataset, the analysis of scientific interest (or `substantive model') is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}$-values are obtained. These $m$ values are combined into a single pooled estimate $\bar{Q}$.}} %\footnotemark}
\end{figure}

%\footnotetext{Figure .}

% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
% "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample. We define the response indicator R as an n x p 0-1 matrix. The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ... , n and j = 1 , ... , p . If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ... The observed data are collectively denoted by Y o b s . The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 . When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}
% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing. The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | Y o b s , $\psi$ ) (2.2) so the missingness probability may depend on observed information, including any design factors. Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself. \cite[par.~2.2]{buur18}

% Also introduce: 
% 
% - Terminology (MCAR, MAR, MNAR)?
% 
% - Blue points are observed, the red points are imputed?

@@

There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}. Some default techniques in \pkg{mice} might not yield converged states at all \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully. The diagnostic evaluation of convergence is preferred over the current practice---visually inspecting imputation chains for signs of non-convergence---since the latter may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}. %Diagnostic evaluation of algorithmic convergence does not share thisApplied researchers should be facilitated in evaluatingClear guidelines for the diagnostic evaluation of convergence would be preferred.

%In that respect is the current evaluation practice insufficient. That is, the  recommendation is to visually inspect imputation chains for signs of non-convergence, which may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}. Quantitative evaluation of the convergence of MI algorithms would be preferred.
The application of convergence diagnostics to MI methods has not been systematically studied \citep{buur18}. The measures that are available for diagnosing convergence of iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., \pkg{mice} algorithms or Gibbs samplers) generally evaluate signs of non-convergence \citep{hoff09}. Non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling). The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ \citep[a.k.a. `Gelman-Rubin statistic';][]{gelm92}\footnotemark. The stationarity component may be assessed through autocorrelations. Other convergence diagnostics are outside of the scope of this study.
%(Several convergence diagnostics have been proposed for evaluating convergence of iterative algorithmic procedures

@@

% The multiple imputation algorithm in \pkg{mice} is a special case of Markov chain Monte Carlo (MCMC) methods---a framework of iterative algorithmic procedures. Within this framework, several convergence diagnostics have been proposed, but the application of these measures on MI algorithms has not been systematically studied \citep{buur18}.
% % For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data. This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}). 
% %%%%%%%
% %This section provides a brief review of: \textbf{[INMIDDELS INCORRECT]}% briefly reviews?
% %1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms. 
% 
% Loosely speaking, MCMC algorithms aim to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Gibbs samplers, or the distribution of missing values in MI algorithms). Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}. Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}. In practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}. The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. Stationarity refers to the absence of trending across iterations within chains. %\textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}  
% %The aim of MCMC procedures is to sample from an unknown (theoretical) distribution. The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.
% 
% % Assessing convergence is difficult. We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}. % In general: "In general, you cannot know for sure if your chain has converged. But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
% % More importantly, in MCMC algorithms like MICE, "what is produced by the algorithm at convergence is not a single number or even a distribution, but rather a sample from a distribution" \citep[p.~883]{cowl96}. "automated convergence monitoring (as by a machine) is unsafe and should be avoided" \citep[p.~902]{cowl96}.
% %Convergence has two components: `mixing' and `stationarity' \citep[p.~284]{gelm13}. Mixing refers to the variation between chains of samples. Stationarity has to do with stability of samples within chains. Ideally, both components will be satisfied: there are no trends within chains, and the chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. 
% 
% \textbf{[COULD BE SHORTER! Maybe in footnote??]} Most convergence diagnostics target either of the two components. The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$; ][]{scha97, gelm13}, numeric standard error \citep[or `MC error'; ][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine the effect of trending within chains. The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[a.k.a. `Gelman-Rubin statistic';][]{gelm92}. With an adapted version of $\widehat{R}$, proposed by \cite{veht19}, we might also evaluate the stationarity component of convergence. This would make $\widehat{R}$ a general convergence diagnostic. The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated. Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by \cite[p.~898]{cowl96}.
% %Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with autocorrelation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
% Other convergence diagnostics are outside the scope of this study.\footnotemark 
% \footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). \cite{buur18} proposed to implement multivariable evaluation through eigenvalue decomposition \cite{mack03}. This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$ and autocorrelation.
% 
% %Diagnose non-convergence in the mixing sense: $\widehat{R}$. Diagnose non-convergence in the stability sense: autocorrelation (AC). Also, it is possible to compute MC errors, the Geweke statistic \textbf{\citep{XYZ}}, or to determine the appropriate number of iterations for convergence by employing \textbf{Raftery and Lewis's (1992)} method. All of these methods evaluate the convergence of univariate scalars of interest (e.g., chain mean or chain variance). The convergence of multivariable statistics (i.e. relations between such scalars) may be evaluated via the eigenvector decomposition method proposed by McKay \citep{XYZ}. These four methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$ and autocorrelation.}
% 
% % "In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points. Do these properties hold for the MICE algorithm? Irreducibility is generally not a problem since the user has large control over the interesting parts of the state space. This flexibility is actually the main rationale for FCS instead of a joint model. Periodicity is a potential problem, and can arise in the situation where imputation models are clearly inconsistent. A rather artificial example of an oscillatory behavior occurs when Y 1 is imputed by Y 2 $\beta$ + $\epsilon$  1 and Y 2 is imputed by - Y 1 $\beta$ + $\epsilon$  2 for some fixed, nonzero $\beta$ . The sampler will oscillate between two qualitatively different states, so the correlation between Y 1 and Y 2 after imputing Y 1 will differ from that after imputing Y 2 . In general, we would like the statistical inferences to be independent of the stopping point. A way to diagnose the ping-pong problem, or order effect, is to stop the chain at different points. The stopping point should not affect the statistical inferences. The addition of noise to create imputations is a safeguard against periodicity, and allows the sampler to "break out" more easily. Non-recurrence may also be a potential difficulty, manifesting itself as explosive or non-stationary behavior. For example, if imputations are made by deterministic functions, the Markov chain may lock up. Such cases can sometimes be diagnosed from the trace lines of the sampler. See Section 6.5.2 for an example. As long as the parameters of imputation models are estimated from the data, non-recurrence is mild or absent. The required properties of the MCMC method can be translated into conditions on the eigenvalues of the matrix of transition probabilities (MacKay 2003, 372-73). The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
%

% The convergence diagnostics under consideration are $\widehat{R}$ and autocorrelation. Perfect convergence is reached when the variance between imputation chains is equivalent to the variance within chains ($\widehat{R} = 1$), and there is no dependency between subsequent iterations of imputation chains ($AC = 0$). 
%
%%%%%%%%%%%%%%%%%
%As potential scale reduction factor, $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \citep[$t \to \infty$;][]{gelm92}. If there is no added value compared to infinitely many iterations, the chains have converged. 
%%%%%%%%%%%%%%%%%%
%
%$\textbf{Explain the measures that I'm investigating here. R hat is ... autocorrelation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
%Explain that there is no baseline to compare the measures with. The only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
%
%The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \cite{gelm92}. That then informs us about how dependent the chains are on the starting values. If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word). 
%$\widehat{R}$ is then equal to one. 
%
%We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest.

To define $\widehat{R}$, we follow notation by \cite[p.~5]{veht19}. % The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}
Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance). For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

%\textbf{REPHRASE: In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains. For each scalar summary of interest $\theta$, we compute $B$ and $W$, the between- and within-chain variances:}

@@

%The proportion of within chain variance $W$ against the weighted average of $B$ and $W$, $\widehat{\operatorname{var}}^{+}$ tells us how much variance of $\theta$ could be shrunken down if $N\to\infty$. $\widehat{\operatorname{var}}^{+}$ is computed as follows: %over-estimates the marginal posterior variance of the estimand, $\theta$.

% \begin{equation*}
% 
% \end{equation*}

%The potential scale reduction factor $\widehat{R}$ is obtained as:
We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}. This interpretation assumes that chains are `over-dispersed' at $t=1$, and reach convergence as $T \to \infty$. Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other. When all chains sample independent of their initial values, the mixing component of convergence is satisfied, and $\widehat{R}$-values will be close to one. High $\widehat{R}$-values thus indicate non-convergence. %The iteration at which chains are independent of their initial values is the point where mixing occurs. This is diagnosed with $\widehat{R}$-values close to one. %If there is no added value 
%High $\widehat{R}$-values thus indicate slow mixing of chains, which is a sign of non-convergence. %, e.g., because chains may depend on their initial value. %are of dependence across chains on initial values. Once there is no dependence on initial values anymore, there is no added value of continuing iterations. \textbf{[MOGELIJK OVERBODIGE ZIN HIER]} The measure then diagnoses algorithmic convergence ($\widehat{R} = 1$). 
%If there is no added value compared to infinitely many iterations, the chains have converged. 
The conventionally acceptable threshold for convergence was $\widehat{R} < 1.2$ \citep{gelm92}. More recently, \cite{veht19} proposed a more stringent threshold of $\widehat{R} < 1.01$.

@@

We can interpret %the sign and magnitude of ... Any non-zero $AC$-value ($-1 \leq AC \leq 1$) . $AC = 0$ indicates independence between $\theta$-values.
$AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains. Positive $AC$-values, however, indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. % (i.e., similarity between $\theta$-values of subsequent iterations), which may lead to trending. Negative $AC$-values occur when the $\theta$-values of subsequent iterations diverge from one-another. Dependence in this sense does not threaten the stationarity component of convergence. %that we aim to evaluate with $AC$. 
On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one-another, which may increase the variance of $\theta$ and speed up convergence.  %have also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}. Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$. 
As convergence diagnostic, the interest is therefore in positive $AC$-values.%\footnotemark 
\footnote{Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.}

@@

%, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances).
%Non-recurrence can be evaluated with autocorrelation. autocorrelation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. 
%A high autocorrelation indicates dependence within chains. The magnitude of the AC can be interpreted qualitatively or quantitatively. Quantitative evaluation of the AC entails comparing the observed ACs to the critical values of a two-tailed 95\% confidence interval, divided by the square root of the number of iterations. %. critical value corresponding to the $z-$~value standard normal distribution with \textbf{XYZ} degrees of freedom.
%  AC_lim <- qnorm(.975) / sqrt(maxit)

>>>>

%% -- Set-up document -------------------------------------------------------------
%%
%%
%%
%%

@@

%% more packages 
\usepackage{multirow, amsmath, tikz} % for table, equations, and flow chart figure (respectively)
\usetikzlibrary{fit, positioning} % for flow chart figure

@@

%% - \Abstract{} almost as usual
\Abstract{
This note investigates how algorithmic convergence of multiple imputation procedures---methodology to circumvent the ubiquitous problem of missing data---may be diagnosed. We conclude that potential scale reduction factor $\widehat{R}$ and autocorrelation may be appropriate convergence diagnostics, but their performance differs from conventional applications to iterative algorithmic procedures. These results will be implemented into an interactive framework for the evaluation of multiple imputation methods that is under development as Research Master's thesis.}

@@

%% -- Introduction -------------------------------------------------------------
%%
%%
%%
%%

\section{Introduction} \label{sec:intro}

@@

To obtain valid inferences with MI, the variability between imputations should be properly represented \citep{rubin87, buur18}. If this variability is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. Over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power. Since both of these situations are undesirable, imputations and their variability should be evaluated. Evaluation measures, however, are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}. 
The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods. These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation, which will aid applied researchers in drawing valid inference from incomplete datasets.

This note provides the theoretical foundation towards the diagnostic evaluation of %the convergence of 
multiple imputation algorithms. For reasons of brevity, we only focus on the MI algorithm implemented in \pkg{mice} \citep{mice}. 
The convergence properties of this MI algorithm are investigated through model-based simulation\footnotemark.%The results of this simulation study are guidelines for assessing convergence of MI algorithms. These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. 
\footnotetext{All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.} 
We will evaluate how convergence of the MI algorithm can be diagnosed.

%% -- Terminology ---------------------------------------------------------------
\subsection{Terminology} \label{sec:terms}

%The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problem. 
This note follows notation and conventions of \pkg{mice} \citep{mice}. % to suit the \pkg{mice} environment, see \cite[p.~4]{mice}. %Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.
Basic familiarity with MI methodology is assumed.\footnote{For the theoretical foundation of MI, see \cite{rubin87}. For an accessible and comprehensive introduction to MI from an applied perspective, see e.g. \cite{buur18}.}

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted by $Y_{obs}$, and will be referred to as `incomplete' or `observed' data. The missing part of $Y$ is denoted by $Y_{mis}$. Which parts of $Y$ are missing is determined by the `missingness mechanism'.  
This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$ \citep{rubin87}.

@@

\begin{figure}
\label{fig:steps}
\centering
	\large{\begin{tikzpicture}
	\tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 20mm]
\tikzstyle{connect}=[-latex, thick]
\node[main, fill = white!100] (data) [label=left:Incomplete data] { };
\node[main] (mids) [right=of data] { };
\node[main] (mids2) [above=10mm of mids] { };
\node[main] (mids3) [below=10mm of mids,label=below:Imputed data] { };
\node[main] (mira) [right=of mids] {};
\node[main] (mira2) [above=10mm of mira] { };
\node[main] (mira3) [below=10mm of mira,label=below:Analysis results] { };
\node[main, fill = white!100] (mipo) [right=of mira,label=right:Pooled result] { };
\path (data) edge [connect] (mids)
      (data) edge [connect] (mids2)
      (data) edge [connect] (mids3)
      (mids) edge [connect] (mira)
      (mids2) edge [connect] (mira2)
      (mids3) edge [connect] (mira3)
		  (mira) edge [connect] (mipo)
		  (mira2) edge [connect] (mipo)
		  (mira3) edge [connect] (mipo);
\end{tikzpicture}}
\caption{Scheme of the main steps in multiple imputation \citep[$m = 3$; adapted from][\S~1.4.1]{buur18}. {\footnotesize Missing data in dataset $Y$ is `imputed' (i.e., filled in) $m$ times. The imputed data is combined with the observed data ($Y_{obs}$) to create $m$ completed datasets. On each completed dataset, the analysis of scientific interest (or `substantive model') is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $m$ separate $\hat{Q}$-values are obtained. These $m$ values are combined into a single pooled estimate $\bar{Q}$.}} 
\end{figure}

@@

There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}. Some default techniques in \pkg{mice} might not yield converged states at all \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully. The diagnostic evaluation of convergence is preferred over the current practice---visually inspecting imputation chains for signs of non-convergence---since the latter may be challenging to the untrained eye \citep[\S~6.5.2]{buur18}.

The application of convergence diagnostics to MI methods has not been systematically studied \citep{buur18}. The measures that are available for diagnosing convergence of iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., \pkg{mice} algorithms or Gibbs samplers) generally evaluate signs of non-convergence \citep{hoff09}. Non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling). 
% Most convergence diagnostics target either of the two components. The stationarity component of convergence may be evaluated with autocorrelation \citep[$AC$; ][]{scha97, gelm13}, numeric standard error \citep[or `MC error'; ][]{gewe92}, and Raftery and Lewis's (\citeyear{raft91}) procedure to determine the effect of trending within chains. 
%\footnote{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). \cite{buur18} proposed to implement multivariable evaluation through eigenvalue decomposition \cite{mack03}. This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{mcka03}.} 
% The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[a.k.a. `Gelman-Rubin statistic';][]{gelm92}. With an adapted version of $\widehat{R}$, proposed by \cite{veht19}, we might also evaluate the stationarity component of convergence. This would make $\widehat{R}$ a general convergence diagnostic. The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated. Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by \cite[p.~898]{cowl96}.
The mixing component of convergence may be evaluated with the potential scale reduction factor, $\widehat{R}$ \citep[a.k.a. `Gelman-Rubin statistic';][]{gelm92}\footnotemark. The stationarity component may be assessed through autocorrelations. Other convergence diagnostics are outside of the scope of this study.

@@

To define $\widehat{R}$, we follow notation by \cite[p.~5]{veht19}. Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance). For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}. This interpretation assumes that chains are `over-dispersed' at $t=1$, and reach convergence as $T \to \infty$. Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other. When all chains sample independent of their initial values, the mixing component of convergence is satisfied, and $\widehat{R}$-values will be close to one. High $\widehat{R}$-values thus indicate non-convergence. The conventionally acceptable threshold for convergence was $\widehat{R} < 1.2$ \citep{gelm92}. More recently, \cite{veht19} proposed a more stringent threshold of $\widehat{R} < 1.01$.

@@

We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains. Positive $AC$-values, however, indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one-another, which may increase the variance of $\theta$ and speed up convergence. As convergence diagnostic, the interest is therefore in positive $AC$-values. 
\footnote{Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.}

@@

%% -- Hypothesis -------------------------------------------------------------
\subsection{Simulation Hypothesis} \label{sec:hypothesis}
